A Mechanised Library for Generic (Co)Programming

### 2024-02-16 paper/paper.tex
Paper

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect.
\begin{minted}{Coq}
Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Set) :=
  { shape : Shape; contents : {p | valid shape pos} -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
omit here the proofs that these container extensions are isomorphic to
polynomials, but this proof is straightforward. To illustrate how the identity,
constant, sum and product container extensions work, consider the
tree-generating functor (i.e. the functor whose fixpoint is a binary tree):
$F\; X = \mathbb{N} + X \times X$.

@@

Definition Cs : Type := App F S2.
Definition Cp : Type := P1 * P2.
Instance Comp : Cont Cs Cp := 
{ valid :=  
    fun (s : App F S2) (p : P1 * P2) =>
        match valid (shape s) (fst p) 
        with
        | true  => fun H =>
            valid (cont s {fst p | H}) 
                  (snd p)
        | false => fun _ => false
        end eq_refl }.
  \end{minted}
\end{minipage}
  \caption{Polynomials as Container Extensions. In order: identity (L4-7),
  constant (L9-13), product (L15-22), sum (L23-31), composition (L33-44).}
  \label{fig:poly}
\end{figure}

@@

The equality of container extensions needs to be extensional, since the
contents are obtained by applying a function. We want to equate container
extensions that have the same shape, and that, for equal shape and position,
they return the same element. 
\begin{minted}{Coq}
  { shape : Shape; contents : {p | valid shape pos} -> X }.
\end{minted}

@@

Reasoning about the equality of containers requires reasoning about the
equality of the contained elements. But the contained elements are given by
function \mintinline{Coq}{contents}. This means that container equality needs
to be extensional, if we want to avoid the functional extensionality axiom.
Furthermore, the equality of functions also needs to be extensional. To deal
with the fact that different types will use different notions of equality, we
restrict to types with an associated equivalence relation, and \emph{proper
morphisms}, i.e. morphisms that respect the necessary equivalence relations.

@@

 Structure morph :=
   MkMorph
     { app :> A -> B;
       app_eq : forall x y, x =e y -> app x =e app y
     }.
Arguments morph A {eA} B {eB}.
Notation "x ~> y" := (morph x y).
\end{minted}
\dcas{Simplify the above}

@@

By default, we associate every type with Coq's propositional equality, but allow
the use of other forms of equality. In most cases, we will use either
propositional equality, extensional equality, or a mixture of both, e.g. for
container extensions, we use an arbitrary equality for shapes, extensional
equality for the contained elements, but Coq's propositional equality for
positions:

\begin{minted}{Coq}
  Inductive AppR `{F : Cont Sh P} (X : Type) {e : equiv X}
           (x y : App F X) : Prop :=
  | AppR_ext
      (Es : shape x =e shape y)
      (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{minted}
The reason for this is to avoid a circularity in the definitions: arbitrary
equivalences for positions would force us to define the contents as morphisms
from positions to the contained elements, but that would as well force us to
redefine container extensions so that they are only applicable to types with
arbitrary equalities and morphisms.  We will see in the following section why
this is problematic for defining initial algebras of container extensions.
\dcas{I am actually not sure of what I wrote above, and it may be something I
simply overlooked. I will look into the Coq code and try to generalise it if I
have the time}

We associate some types with their standard equivalences:
\begin{minted}{Coq}
Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
Instance prop_eq : equiv Prop.
Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
\end{minted}

>>>>

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{minted}{Coq}
Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Set) :=
  { shape : Shape; contents : PosOf C shape -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
omit here the proofs that these container extensions are isomorphic to
polynomials, but this proof is straightforward. To illustrate how the identity,
constant, sum and product container extensions work, consider the
tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
$F\; X = \mathbb{N} + X \times X$.

@@

Definition Cs : Type := App F S2.
Definition Cp : Type := P1 * P2.
Instance Comp : Cont Cs Cp := 
{ valid :=  
    fun (s : App F S2) (p : P1 * P2) =>
        match valid (shape s) (fst p) 
        with
        | true  => fun H =>
            valid (cont s {fst p | H}) 
                  (snd p)
        | false => fun _ => false
        end eq_refl }.
  \end{minted}
\end{minipage}
  \caption{Polynomials as Container Extensions (simplified). In order: identity
  (L4-7), constant (L9-13), product (L15-22), sum (L23-31), composition
  (L33-44).}
  \label{fig:poly}
\end{figure}

@@

The equality of container extensions needs to be extensional, since the
contents are obtained by applying a function. We want to equate container
extensions that have the same shape, and that, for equal shape and position,
they return the same element. To avoid the use of functional extensionality
axioms, we capture this relation with the following inductive proposition in
Coq:
\begin{minted}{Coq}
Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
| AppR_ext (Es : shape x = shape y)
           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
\end{minted}
Note that  we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is
trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \mintinline{Coq}{App F A -> B} (or \mintinline{Coq}{B -> App F A}). 
We want the following kind of properties to hold: 
\begin{minted}{Coq}
forall (x y : App C A) (f : App C A -> B), AppR x y -> f x = f y
\end{minted}
Moreover, we want this to be true for any \mintinline{Coq}{B}, and equality.
We address this in the following section.

@@

The need to reason about different forms of equality forces us to couple every
type with an associated equivalence relation. Furthermore, we will only allow
to reason about the equality of \emph{proper morphisms}, i.e. morphisms that
respect the necessary equivalence relations.

@@

 Structure morph :=
   MkMorph
     { app :> A -> B;
       app_eq : forall x y, x =e y -> app x =e app y
     }.
Arguments morph A {eA} B {eB}.
Notation "x ~> y" := (morph x y).
\end{minted}

@@

By default, we associate every type with Coq's propositional equality, but allow
the use of other forms of equality. In most cases, we will use either
propositional equality, extensional equality, or a mixture of both, e.g. for
container extensions, we use an arbitrary equality for shapes, extensional
equality for the contained elements, but Coq's propositional equality for
positions:
\begin{minted}{Coq}
  Inductive AppR `{Cont Sh P} `{equiv X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{minted}
Notice the use of \mintinline{Coq}{=e} for both shapes, and the continuations
of container extensions. 
The reason to use Coq's propositional equality for positions is to avoid a
circularity in the definitions: arbitrary equivalences for positions would
force us to define the contents as morphisms from positions to the contained
elements, but that would as well force us to redefine container extensions so
that they are only applicable to types with arbitrary equalities and morphisms.
We will see in the following section why this is problematic for defining
initial algebras of container extensions.
\dcas{I am actually not sure of what I wrote above, and it may be something I
simply overlooked. I will look into the Coq code and try to generalise it if I
have the time}

\subsubsection{Polynomial Types, Equivalences, and Morphisms}
We associate some types with their standard equivalences:
\begin{minted}{Coq}
Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
Instance prop_eq : equiv Prop.
Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
\end{minted}

\subsubsection{Isomorphisms with Container Extensions}
